%% $Id: acknowledgements.tex,v 1.1 2006/01/30 21:12:08 pwaddell Exp $

{\bf \large Acknowledgments}

The project described in this report has been funded principally
by the Puget Sound Regional Council. It was also supported by
grants from the National Science Foundation Grants CMS-9818378,
EIA-0121326 and EIA-0090832, which have supported the development
of UrbanSim.

Numerous people have contributed substantially to the development
of this report and the analysis it contains.  In particular, we
wish to acknowledge the support of Mark Simonson and Larry Blain
at the Puget Sound Regional Council, who coordinated the Technical
Advisory Committee. Considerable assistance was also provided by
Kevin Murphy, Jeff Frkonja, Carol Naito, Jerry Harless, Andy
Norton, Kristen Koch, Neil Kilgren, and others at the PSRC.
Members of the Technical Advisory Committee, listed below, were
instrumental in improving the database and providing valuable
feedback in this effort.  At the Center for Urban Simulation and
Policy Analysis, Chris Peak and Peter Caballero carried out much
of the database development effort, with considerable assistance
from Jack Kim, who analyzed non-residential real estate, Kapena
Pflum, who assisted in documentation and in verification of
employment, Liming Wang, who developed scripts for data
processing, Bjorn-Freeman Benson, Rob Duisberg and David Socha,
who developed key software components to facilitate the data
analysis and processing.

{\bf \large Center for Urban Simulation and Policy Analysis}

The Center for Urban Simulation and Policy Analysis (CUSPA) at the
University of Washington brings advanced research and information
technology to bear on complex social, economic, and environmental
problems in metropolitan areas.  CUSPA is committed to the
development and pursuit of an integrated long-term research agenda
to help understand and anticipate the consequences of policy
choices regarding transportation, land use, housing, community and
economic development, and environmental quality, and to enable
more effective development, implementation, and evaluation of
policies. CUSPA is equally committed to making this research
widely accessible to policymakers and communities through the
development of information technology for advanced simulation
modeling, visualization, and decision support to facilitate
informed and democratic community policy decisions that involve
uncertainty, complex interdependencies, and conflicts over values
and strategies.

CUSPA is structured as an interdisciplinary research center at the
University of Washington, drawing together expertise from across
the campus to address common challenges.  It engages in strategic
partnerships with public agencies to facilitate linking research,
application, and education.  In short, CUSPA attempts to bring the
best available research to bear on the most pressing urban and
environmental problems faced by metropolitan regions in the U.S.
and elsewhere.

\begin{tabbing}
Director: Director: Director: \= \kill Director:\> Paul Waddell,
Evans School of Public Affairs \\
Associate Director:\> Alan Borning, Computer Science and Engineering \\
Steering Committee:\> Marina Alberti, Urban Design and Planning \\
               \> Batya Friedman, Information School \\
\>Scott Rutherford, Civil and Environmental Engineering \\\\
Center Home Page: \> {\bf www.cuspa.washington.edu} \\
UrbanSim Home Page: \> {\bf www.urbansim.org}
\end{tabbing}

\newpage
{\bf \large Technical Advisory Committee} \\

Travis Black -- Kitsap County \\
Larry Blain -- Puget Sound Regional Council\\
Dan Cardwell -- Pierce County \\
Greg Cioc -- Kitsap County \\
Steve Cohn -- City of Bellevue \\
Chandler Felt -- King County \\
Lauren Giboney -- Snohomish County \\
Michael Hubner -- The Suburban Cities Association of King County \\
Tim Koss --Snohomish County \\
Kevin Murphy -- Puget Sound Regional Council\\
Scott Murphy -- Kitsap County \\
Carol Naito -- Puget Sound Regional Council\\
Jennifer Pettyjohn -- City of Seattle \\
Shawn Phelps -- Pierce County \\
Jason Rice -- Kitsap County \\
Mark Simonson -- Puget Sound Regional Council\\
Steve Toy -- Snohomish County \\
Lish Whitson -- City of Seattle \\
