\subsection{System Security}
There are many situations in which there may be concerns related to restricting access to data or functions to different users and/or machines.  We propose to exploit existing infrastructure wherever possible to address security requirements.  Several security points are available using existing system functionality:


\begin{itemize}
\item Database Access: Access to the Geodatabase may be protected at the user or group level, and can be applied to databases or possibly to the table level, for managing read and write access to data.  This would provide a lightweight mechanism for managing user access, for example by restricting which users or groups have access to a database containing the database containing the core data for the AZ-SMART application.  More fine-grained control would also be possibe, for example protecting employment data from read access outside a set of users within the agency that have been cleared for this access.

\item Login Access: Each machine that has AZ-SMART installed will have login access managament, which can be used as a simple but effective way to prevent unauthorized access to the AZ-SMART system.

\item Machine Access: A machine may be enabled for access by setting environment variables that can store username and password information, and this can be loaded by the az-smart system as needed for security enforcement.  A generic az-smart user could be defined, or a group account, or an individual user account and a password assigned to each.  These usernames and passwords could be stored in an az-smart-security database that only an AZ-SMART Administrator is enabled to access and maintain.

\end{itemize}
