%% $Id: acknowledgements.tex,v 1.2 2004/07/02 20:57:36 pwaddell Exp $

{\bf \large Acknowledgments}

The development of OPUS and UrbanSim has been supported by
grants from the National Science Foundation Grants CMS-9818378,
EIA-0090832, EIA-0121326, IIS-0534094, and IIS-0705898,
and by the Maricopa Association
of Governments, the Puget Sound Regional Council, and
the contributions of users. Numerous people have contributed 
substantially to the development of this project. See the
UrbanSim People page for current and previous contributors,
at {\bf www.urbansim.org/people}.   

{\bf \large Center for Urban Simulation and Policy Analysis}

The Center for Urban Simulation and Policy Analysis (CUSPA) at the
University of Washington brings advanced research and information
technology to bear on complex social, economic, and environmental
problems in metropolitan areas.  CUSPA is committed to the
development and pursuit of an integrated long-term research agenda
to help understand and anticipate the consequences of policy
choices regarding transportation, land use, housing, community and
economic development, and environmental quality, and to enable
more effective development, implementation, and evaluation of
policies. CUSPA is equally committed to making this research
widely accessible to policymakers and communities through the
development of information technology for advanced simulation
modeling, visualization, and decision support to facilitate
informed and democratic community policy decisions that involve
uncertainty, complex interdependencies, and conflicts over values
and strategies.

CUSPA is structured as an interdisciplinary research center at the
University of Washington, drawing together expertise from across
the campus to address common challenges.  It engages in strategic
partnerships with public agencies to facilitate linking research,
application, and education.  In short, CUSPA attempts to bring the
best available research to bear on the most pressing urban and
environmental problems faced by metropolitan regions in the U.S.
and elsewhere.

\begin{tabbing}
Director: Director: Director: \= \kill Director:\> Paul Waddell,
Evans School of Public Affairs \\
Co-Director:\> Alan Borning, Computer Science and Engineering \\

%Center Home Page: \> {\bf www.cuspa.washington.edu} \\
UrbanSim Home Page: \> {\bf www.urbansim.org} 
\end{tabbing}

