% Copyright (c) 2005-2009 Center for Urban Simulation and Policy Analysis,
% University of Washington.  Permission is granted to copy, distribute and/or
% modify this document under the terms of the GNU Free Documentation License,
% Version 1.2 or any later version published by the Free Software Foundation;
% with no Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts.
% A copy of the license is included in the section entitled "GNU Free
% Documentation License".

\chapter{The \package{urbansim} Opus Package}

\emph{NOTE: this is preliminary draft documentation of the UrbanSim model system, incorporating
plans for extensions of the system to address the needs of the AZ-SMART model system and 
the PSRC model system for a parcel-based version of the system.  It is not fully implemented in 
code at this point.  Once code and
documentation are consistent, this will be folded into the main UrbanSim documentation.  Separate, 
brief documentation will be generated for the configuration of these models for specific
applications like PSRC and AZ-SMART.}

\section{Introduction}

UrbanSim is implemented (as of version 4) as a package in the Open Platform for Urban Simulation (Opus).
It is a general and customizable package for creating models and model systems, estimating model
parameters, creating scenarios for alternative policy assumptions, running the model system on these
scenarios, and evaluating them.  In the current version of UrbanSim, support has been added for 
generalizing the use of different kinds of geographic units.  Previously the system has been implemented
with a focus on the use of gridcell geography only.  Now it supports implementing models and model
systems using alternatives such as parcels (and buildings), or zones, which could represent any spatial
unit of analysis that aggregates from parcels or gridcells.  These geographies can now also be mixed -- 
allowing zonal variables to be computed and used in a gridcell or parcel model, or gridcell variables to be
used in a parcel model, for example.  As a result of this new flexibility, the documentation and the urbansim 
Opus package have been reorganized.

The urbansim package makes extensive use of functionality in the opus_core package, and provides additional
classes to implement datasets, models, and model systems.  The package has been re-organized into the 
following structure (\emph{note: this has not yet been completed, and the following is only for illustration purposes}):

urbansim
\begin{itemize}
	\item{data}
		\begin{itemize}
			\item{building}
			\item{business}
			\item{development_constraints}
			\item{development_project_proposal}
			\item{development_site}
			\item{development_template}
			\item{gridcell}
			\item{household}
			\item{job}
			\item{parcel}
			\item{person}
			\item{planned_land_use}
			\item{region}
			\item{travel_data}
			\item{zone}
		\end{itemize}
	\item{models}
		\begin{itemize}
			\item{development_project_location_choice_model|}
			\item{development_project_proposal_choice_model}
			\item{employment_location_choice_model}
			\item{employment_relocation_choice_model}
			\item{employment_transition_model}
			\item{household_location_choice_model}
			\item{household_relocation_choice_model}
			\item{household_transition_model}
			\item{land_price_model}
			\item{real_estate_price_model}
		\end{itemize}
\end{itemize}

\emph{Note: the datasets and models reflect available modules, not all of which would be used in a
particular model application.}

\section{Datasets}
\label{sec:urbansim-datasets}
 
%
All datasets defined in \package{urbansim} are implemented as children of the
Opus \package{opus_core} class \class{Dataset} described in
Section~\ref{sec:opus-core-datasets}. Each dataset sets default values for several
class properties, such as a name of the unique identifier (\verb|id_name|),
dataset name (\verb|dataset_name|), \verb|in_table_name| and
\verb|out_table_name|. 

The \package{urbansim} datasets are defined in \verb|urbansim.datasets|. 
We use the following naming convention: For module name we use the dataset 
name in lower case in singular form, where single words are connected by
'_', and ending with ``_dataset''.
For class name we capitalize the first letters in each word of the dataset 
name, use singular form and add 'Dataset' at the end. For example, a dataset for
development projects is defined in class \class{DevelopmentProjectDataset}
implemented in the module \file{development_project_dataset.py}. For interaction
sets, we connect the two dataset names in the same way, but with 
an 'x' in the module name and an 'X' in the class name. For example, an
interaction set of development projects and gridcells is defined in
\class{DevelopmentXProjectGridcellDataset} implemented in
\file{development_project_x_gridcell_dataset.py}.

\subsection{Predefined Datasets}
 
%
Table~\ref{tab:urbansim-datasets} lists dataset classes that are predefined in
\package{urbansim} (in alphabetical order), including the default value for
\verb|dataset_name|, \verb|in_table_name|, \verb|id_name| and name of the
module in which the class is implemented (excluding '.py'). In most cases they
correspond to database tables described in
Chapter~\ref{chapter:urbansim-database-tables}. The corresponding table name
is the value of \verb|in_table_name|.

%begin{latexonly}
%end{latexonly}
\begin{table}
\begin{center}
\begin{tabular}{|l||l||l|}\hline
\multicolumn{1}{|c||}{dataset_name} & \multicolumn{1}{c||}{in_table_name (default)}&
\multicolumn{1}{c|}{id_name (default)} \\\hline\hline
%
city* & cities & city_id 
\\
building & buildings & building_id 
\\
building_type & building_types & building_type_id 
\\
control_total (households) & annual_household_control_totals & year 
\\ 
control_total (jobs) & annual_employment_control_totals & year, sector_id 
\\ 
county* & counties & county_id 
\\
development_constraint & development_constraints & constraint_id 
\\ 
development_event & development_events_exogenous & grid_id, scheduled_year 
\\
development_event_history & development_event_history & grid_id, scheduled_year 
\\
development_group & development_type_groups & group_id 
\\
development_project & -- & project_id 
\\
development_project_proposal & -- & development_project_proposal_id
\\
development_site* & development_sites & development_site_id
\\
development_type & development_types, & development_type_id 
\\
& development_type_group_definitions & 
\\
employment_sector & sector_id & employment_sectors 
\\
employment_sector_group & employment_adhoc_sector_groups & group_id 
\\
faz* & fazes & faz_id 
\\
fazdistrict* & -- & fazdistrict_id 
\\
gridcell* & gridcells & grid_id 
\\
household_characteristic & household_characteristics_for_ht & --
\\
household & households & household_id
\\
job_building_type & job_building_types & id 
\\
job & jobs & job_id 
\\
large_area* & large_areas & large_area_id 
\\
neighborhood* & neighborhoods & neighborhood_id 
\\
parcel* & parcels & parcel_id
\\
planned_development & planned_developments & planned_development_id
\\
plan_type & plan_types & plan_type_id 
\\
plan_type_group & plan_type_groups & group_id 
\\
race & race_names & race_id 
\\
rate (households) & annual_relocation_rates_for_households & age_min, income_min 
\\
rate (jobs) & annual_relocation_rates_for_jobs & sector_id 
\\
region & regions & region_id 
\\
ring & rings & ring_id 
\\
target_vacancy & target_vacancies & year
\\
travel_data & travel_data & from_zone_id, to_zone_id 
\\
vacant_land_and_building_type & & 
\\
zone* & zones & zone_id 
\\\hline
\end{tabular}
\end{center}
\caption{\label{tab:urbansim-datasets}Datasets defined in \package{urbansim}. A dataset
  marked with * is a location set, i.e. it represents a set of locations of
  a specific geographical unit and can be visualized as a two-dimensional image.}
\end{table}
%begin{latexonly}
%end{latexonly}

\section{Variables}

%
Variables predefined in \package{urbansim} are structured according to the
dataset to which they belong. They are implemented in directories of the same
name as \verb|dataset_name| which is placed in the top directory of
\package{urbansim}. For example, all gridcell variables are placed in
``urbansim.gridcell'', all interaction variables of households and
gridcells are placed in ``urbansim.household_x_gridcell''. All variables
are defined as children of the Opus \package{opus_core} class \class{Variable} and
thus they follow the guidelines presented in Section~\ref{sec:opus-variable}.

\section{Models}
\label{sec:urbansim-models}
\modelsindex

\package{urbansim} has a set of predefined models, each
of them implemented as a child of one of the core Opus models
\modelsindex described in Section~\ref{sec:opus-core-models}.  They
are hierarchically structured and a few of them are created using
specific creators.  All
\package{urbansim} models are placed in
``urbansim.models''.


\subsection{Overview}

The models to execute when running UrbanSim are determined by the configuration
(see Section~\ref{sec:model-system-configuration}). Its entry ``models'' is a
list of character strings that identify the models to run in that order.
UrbanSim defines following models, grouped by their use in a gridcell, parcel or zone level model
application: 

\input{parcel-models-list}
\input{parcel-models}
%
\subsection{Choices Classes}
%
\label{sec:urbansim-choices}
This section describes \class{Choices} classes (see
Section~\ref{sec:choices}) that are implemented in \package{urbansim}.

\subsubsection{lottery_choices}
\index{lottery_choices}
%
This class determines choices by taking capacity into account. The
\method{run()} method takes as arguments \verb|probability| which must be a 2D
array (number of observations $\times$ number of (sampled) alternatives) and
\verb|resources| which must contain an entry ``capacity''. This entry should
be a numpy array \numpyindex or list giving capacity of each alternative of the universe.
\verb|resources| can contain an entry ``agent_units'' which is a 1D numpy array \numpyindex
or list giving for each agent the number of units to be removed from capacity
when placing this agent. Default value is 1 for each agent.  \verb|resources|
can also contain an entry ``index'' which is a 1 or 2D array of indices of
alternatives for each agent. If it is a 1D array, it is assumed to be the same
for all agents. If alternatives were previously sampled, it is the index of
the sampled alternatives within the universe. The shape of ``index'' should
correspond to the shape of \verb|probability| or its second dimension if 1D.
If no ``index'' entry is given, it is created for all alternatives given by
``capacity'' and appended to \verb|resources|.

The \method{run()} method invokes the \package{opus_core} module
\class{random_choices_from_index} which returns indices of agent's
choices made independently of one another.  The indices are chosen
values from the entry ``index''. Using capacity, it is then
determined for which choices there is an overflow in agent's
interest. If there is no overflow the method simply returns the
obtained indices. Otherwise, from agents that selected choices with
overflow, a number that corresponds to the overflow is randomly
sampled. Those agents are marked for making a new choice.  The
probability cells that correspond to the not available alternatives
are set to zero and the process is repeated.  The maximum number of
iterations of this process can be controlled by the entry
``lottery_max_iterations'' in \verb|resources| which is by default
3. If there is still overflow after reaching the maximum number of
iterations, in the resulting array there is a value -1 for agents
that couldn't find any choice.


\subsubsection{first_agent_first_choices}
%
\index{first_agent_first_choices}
This class is a child of \class{lottery_choices}. It allows only one agent per
choice. In case of overflow, the first agent keeps the choice, all others must
choose again.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "userguide"
%%% End:
