\chapter{The Data Manager}
The Data Manager has two primary purposes, each reflected in the sub-tabs.  One tab, the Opus Data tab, is for browsing, viewing, and exporting data from the Opus data cache.  The other tab, the Tools tab, is a place for storing and executing various tools provided by the Opus community or tools you have written.

\section{Opus Data Tab}
TODO: describe primary attributes, tables, and "databases"

The Opus Data tab is a file browser that defaults to the folder in which your project creates data.  For instance, if you installed to c:/opus, and you are opening the Eugene sample project at c:/opus/project_configs/eugene_gridcell.xml, the data folder for this project is c:/opus/data/eugene_gridcell.  That is the folder that this view starts at.  Any subfolders and files are displayed in the tree view. (TODO: insert /images/data_manager_opus_data_tab.png)

\subsection{Viewing and Browsing an Opus Data table}
To view an Opus Data table, right-click a data table, then select 'View Dataset'.  (TODO: insert /images/data_manager_opus_data_tab_view_dataset.png) This will bring up a new tab on the right-hand side of the Opus GUI window that will display some summary statistics about the dataset, and a table view of the raw data that can be browsed and sorted by clicking the column name. (TODO: insert /images/data_manager_opus_data_tab_view_dataset_tab.png)

\subsection{Exporting an Opus Data table}
(TODO: insert /images/data_manager_opus_data_tab_export_dataset.png)

\section{Tools Tab}
TODO: general purpose of tools tab

\subsection{Tool Library}
TODO: general purpose of library
TODO: subfolders for organizing tools and how to add them (requires XML right now)
TODO: writing new tools and adding them to the tool library
TODO: executing tools directly from the tool library

\subsection{Tool Sets}
TODO: general purpose of toolsets
TODO: creating new toolsets and adding tools to them
TODO: executing toolsets