%% $Id: data.tex,v 1.5 2001/08/18 00:56:25 borning Exp $

\subsection{Data Import and Export}

On the input side, UrbanSim requires a substantial amount of data
to simulate an urban area, including census, employment, parcel,
zoning, and transportation network data.  We have written a suite
of custom data preparation tools that can convert data to the
formats we require, merge datasets, detect inconsistencies, and in
some cases fill in missing data or correct errors in base year
input data using heuristics.  This input data is generally made
available to UrbanSim in the form of delimited ASCII text files.

On the output side, UrbanSim's output module is responsible for
gathering, aggregating, and exporting data from the Object Store
to a set of external ASCII text files containing requested
extracts or summaries from the Object Store. Currently, the export
model writes output files containing households and jobs by type,
housing units, non-residential square footage, and land and
improvement values, summarized by grid cell and by Traffic
Analysis Zone. These files are then passed on to other systems
(including GIS software, statistical software, spreadsheets, and
the external travel model) for subsequent analysis and graphical
display.  To simplify the software engineering aspects of data
export, we have defined the output module as a model, analogous to
the Demographic Transition Model, the Household Mobility Model,
and so forth (Section \ref{sec:models}).  This allows the Export
Model to be scheduled to run at defined times, querying the Object
Store to take a snapshot of the state of the simulation.
Conceptually, however, it is not a model like the Demographic
Transition Model or others, since it does not represent an agent
in the simulated world, and only performs reads of the Object
Store, not writes.

Finally, in addition to input and output files, the system can
write a snapshot of the current state of the Object Store to an
external file, allowing the system to be restarted quickly from a
given point in the simulation.

% LocalWords:  Exp UrbanSim  UrbanSim's GIS pwaddell
